---
title: "Homework 4"
author: "Manasvi Sundar"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: html_document
---

# Introduction 
## Data used
In this activity, I have used a dataset which records the presence/absence of various mental conditions among 279 patients enrolled in an observational study. The dataset has 256 variables, out of which some indicate presence/absence of mental conditions such as PTSD, depression, psychosis, bipolar disorder, panic disorder and anxiety. For each condition, there are details of further classification and medications taken for the condition. 

## Aim of analysis
My first aim today is to create a summary table to indicate the prevalence of each condition in our study sample. 
My second aim to calculate a mental health score for each patient based on the presence/absence of mental health disorders. 


# Descriptive Analysis
## Loading the required dataset
We now load the datafile which is in csv format by using read.csv() function. Since the dataset has several variables which are not applicable to our aim, we create a subset of only variables we intend to use.

```{r, load-data, results = FALSE }
# loading required data
mydata <- read.csv('Outcomes.csv')

# subsetting only required variables
subdata <- subset(mydata, select = c(patid, SCID_ptsd, SCID_depression, SCID_psychhis, SCID_Bipolar1, SCID_PanicDisorder, SCID_AnxietyGMC))
dim(subdata)
```

We now have a dataset of 279 observations and 7 variables. Since the dataset has already been cleaned, there are no missing values and we do not need to clean the data any further. 

To tabulate the presence/absence of the conditions, we use the table() function for each variable. The variables have been coded as 1 for 'yes' and 0 for 'no'.  

```{r, analysis, results = FALSE}
table(subdata$SCID_ptsd, useNA = "ifany")
table(subdata$SCID_depression, useNA = "ifany")
table(subdata$SCID_psychhis, useNA = "ifany")
table(subdata$SCID_Bipolar1, useNA = "ifany")
table(subdata$SCID_PanicDisorder, useNA = "ifany")
table(subdata$SCID_AnxietyGMC, useNA = "ifany")
```
## Prevalence of disorders in our study sample

Condition         | Present in (%)   | Absent in (%)
------------------|------------------|--------------------
PTSD              | 63 (22.6%)       | 216 (77.4%)
Depression        | 59 (21.1%)       | 220 (78.9%)
Psychosis         | 51 (18.3%)       | 228 (81.7%)
Bipolar Disorder  |	4 (1.4%)         | 275 (98.6%)
Panic Disorder    | 11 (3.9%)        | 268 (96.1%)
Anxiety           | 1 (0.4%)         | 278 (99.6%)


# Calculating a score 

We will not calculate a score for each patient based on the presence/absence of various disorders, since one person may have multiple disorders. 

```{r, calculating-score, results = FALSE}
subdata$score <- subdata$SCID_depression*1 + subdata$SCID_ptsd*3 + subdata$SCID_psychhis*5 + subdata$SCID_Bipolar1*5 + subdata$SCID_PanicDisorder*4 + subdata$SCID_AnxietyGMC*2

table(subdata$score, useNA = "ifany")
mean(subdata$score)
sd(subdata$score)
```

# Conclusion

From the table, it can be concluded that PTSD was most prevalent in the study population (22.6%), followed by depression (21.1%). Anxiety was the least prevalent with only 0.4% of the study population with the condition. 

From our score calculated, the mean score was 2.04 (SD = 3.39). The minimum score was 0 and the maximum score was 14. 


